---
title: "A proposed approach to estimating excess mortality associated with cleft lip / cleft palate"
author: "Mark Myatt"
date: ""
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    latex_engine: xelatex
header_includes: 
  \usepackage{float} 
  \usepackage{setspace} 
  \onehalfspacing
fontsize: 12pt
geometry: margin=2cm
classoption: a4paper
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/"
)

kids <- c(172, 155, 144, 134, 125)
cleft <- c(50, 40, 30, 20, 10)
ages <- c (1, 2, 3, 4, 5)
model <- lm(log(kids) ~ ages)
model

tab <- as.table(kids)
names(tab) <- ages
u5mr <- abs(coef(model)[2]) * 10000 / 365.25
names(u5mr) <- "u5mr"
u5mr

## expected numbers
expected.s0 <- exp(model$coefficients[2] * ages)
expected.p0 <- expected.s0 /  sum(expected.s0)
expected.n0 <- expected.p0 * sum(kids)


cleft.model <- lm(log(cleft) ~ ages)
cleft.model
tabx <- as.table(cleft)
names(tabx) <- ages
cleft.u5mr <- abs(coef(cleft.model)[2]) * 10000 / 365.25
names(cleft.u5mr) <- "cleft.u5mr"
cleft.u5mr

## expected numbers
expected.s1 <- exp(cleft.model$coefficients[2] * ages)
expected.p1 <- expected.s1 /  sum(expected.s1)
expected.n1 <- expected.p1* sum(cleft)
```

This note describes a simple and cheap method that may be useful for the estimating excess mortality associated with cleft lip / cleft palate in settings where reporting of congenital anomalies is incomplete or absent. It is intended as a request for comments.

The method uses:

* Readily available survey data (e.g. census tables, DHS, MICS) or simple survey data that can describe the age-distribution of the general population.

* Data on the ages of found cases from clinical settings (may suffer a selection bias) and from community-based case-finding exercises (e.g. using active and adaptive case-finding).

The method is designed to be very low cost (i.e. compared to a cohort study) and to increase the impact of programs treating cleft lip / cleft palate through the introduction or expansion of community-based case-finding and referral.

Mortality in the general population is estimated from the population age distribution using a simple exponential decay model in which the proportion of the birth cohort surviving at each age follows:

&nbsp;

\begin{equation}
\rho ~ = ~ e ^ {-zt}
\end{equation}

&nbsp;

where $e$ is the base of the natural logarithm (approximately 2.7183), $z$ is the mortality rate, and $t$ is
age. Given an age-distribution (e.g. as shown in Figure 1) we can fit the model:

&nbsp;

\begin{equation}
log_e(n) ~ = ~ \alpha ~ + ~ \beta ~ \times ~ t
\end{equation}

&nbsp;

where $n$ is the count of children in each age group. The absolute value of $\beta$ is an estimate of the mortality rate ($z$) in the general population. For the data shown in Figure 1 this is $z = 0.028205$ which is equivalent to about 0.77 / 10,000 / day. The expected counts for the model:

&nbsp;

\begin{equation}
\rho ~ = ~ e ^ {-0.028205 ~ \times ~ t}
\end{equation}

&nbsp;

are shown in red in Figure 1. We can fit a similar model to the ages of cases of cleft lip / cleft palate found in clinical data and by community-based case-finding. Given the age distribution of cases shown in Figure 2 we estimate $z$ to be 0.36690 which is equivalent to about 10 / 10,000 / day.

Excess mortality can be estimated as the ratio of the two mortality rates:

&nbsp;

\begin{equation}
\text{Morality rate ratio} ~ = ~ \frac{z_\text{cases}}{z_\text{General population}}
\end{equation}

&nbsp;

with the example data this is:

&nbsp;

\begin{equation}
\text{Morality rate ratio} ~ = ~ \frac{0.36690}{0.028205} ~ = ~ 13.01
\end{equation}

&nbsp;

Mortality in cases is about thirteen times higher than the general population. 

A 95% confidence interval could be calculated using the bootstrap.

It is envisaged that data for children and cases aged between 6 months and 5 years (the simple exponential decay model may not hold with older children and adults) will be used and that age will be recorded to best possible accuracy and precision.

\newpage

```{r figure1, echo = FALSE, eval = TRUE, fig.cap = "Example age distribution (0 to 5 years) in the general population", fig.align = "center", fig.retina = 1}
bp <- barplot(tab, xlab = "Age (years)", ylab = "Number of children")
lines(bp[,1], expected.n0, lwd = 2, col = "red")
```

```{r figure2, echo = FALSE, eval = TRUE, fig.cap = "Example age distribution (0 to 5 years) in cases", fig.align = "center", fig.retina = 1}
bp <- barplot(tabx, xlab = "Age (years)", ylab = "Number of children")
lines(bp[,1], expected.n1, lwd = 2, col = "red")
```